---
title: "Machine Learning Course Project"
author: "Phillip"
date: "March 14, 2016"
output: html_document
---

##Background Information
This is a write up for the John Hopkin's Coursea Machine Learning Course Project.  Below is the background infomration for the assignment.  

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Project Goal
The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

##Setting Up Environment 

The following packages will be used during the model building and testing so they are being loaded into the envinvrionment.  I am also setting a seed so this work can be reproduced by others.
```{r}
library(caret)
library(ggplot2)

set.seed(1127)
```


##Datasets 

Next the training and testing datasets are loaded into memory changing all "NA", "#DIV/0!", and "" into "NA" strings to make data cleaning easier.  The datasets could also be downloaded and read them in from the hard disk if you wanted a local copy of the data.  

```{r}
trainURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainURL), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testURL), na.strings=c("NA","#DIV/0!",""))
```


##Data Exploration 
Typically a dataset needs to be split into a training and testing set, however this data has already been split so that step is being skiped.  Next, let's take a look at the size of the training dataset.  
```{r echo=TRUE}
dim(training)
```

The dataset contains 160 variables which would create a rather slow and inefficent model, so the first thing to do is figure out which variables can be removed.  One of the quickest ways is to look at how many acutal values are in each variable and then eliminate the ones missing a good amount of data.  

```{r }
NAcounts <- apply(training, 2, function(x) length(which(!is.na(x))))
table(NAcounts)
```

The output of the table shows that there are 60 variables which have values for every record, while the rest of the variables have values for 406 records or less, so those can be dropped. 

```{r }
training <- training[ , colSums(is.na(training)) == 0]
```

Sixty variables is still a lot so next lets take a look at the names of the varialbes and determine if there are any that can be dropped based on that. 

```{r }
names(training)
```

Based on the names columns 1 thru 5 can be dropped from the dataset as the record number, user name, and timestamp should not impact the model predicition.  

```{r }
training <- training[, -c(1:5)]  
```










